---
title: "DAPPER: Data Augmentation for Private
Posterior Estimation in R"
date: "2023-11-01"
abstract: >
    This paper serves as a reference and introduction on using the dapper R
    package. The goal of this package is to provide some tools for exploring the
    impact of different privacy regimes on a Bayesian analysis. A strength of
    this framework is the ability to target the exact posterior in settings
    where the likelihood is too complex to analytically express.
draft: true
author:  
  - name: Kevin Eng
    affiliation: Rutgers University
    address:
    - Department of Statistics
    - Piscataway, NJ 08854
    url: https://www.britannica.com/animal/quokka
    email: ke157@stat.rutgers.edu
  - name: Jordan A. Awan
    affiliation: Purdue University
    address:
    - Department of Statistics
    - West Lafayette, IN 47907
    url: https://www.britannica.com/animal/quokka
    email: jawan@purdue.edu
  - name: Ruobin Gong
    affiliation: Rutgers University
    address:
    - Department of Statistics
    - Piscataway, NJ 08854
    url: https://www.britannica.com/animal/quokka
    email: ruobin.gong@rutgers.edu
  - name: Nianqiao Phyllis Ju
    affiliation: Purdue University
    address:
    - Department of Statistics
    - West Lafayette, IN 47907
    url: https://www.britannica.com/animal/quokka
    email: nianqiao@purdue.edu
  - name: Vinayak A. Rao
    affiliation: Purdue University
    address:
    - Department of Statistics
    - West Lafayette, IN 47907
    url: https://www.britannica.com/animal/quokka
    email: varao@purdue.edu
type: package
output: 
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(ggplot2)
library(kableExtra)
library(tidyverse)
library(knitr)
```



# Introduction



Differential privacy provides a rigorous framework for protecting 
confidential information. In this framework, privacy is obtained
through the injection of random noise in the data analysis workflow.
It has served as the theoretical foundation for recent advances in privacy technology. Several high profile
examples include Apple (), Google (), Microsoft (), and the 
U.S. Census Bureau ().


There are several approaches for constructing or modifying a data analysis workflow
to provide privacy guarantees, which loosely fall into three categories: direct, query, and dissemination.
The approaches mainly differ in where in the workflow privacy noise is injected. 
In the dissemination setting, data needs to be released to the public, and privacy is maintained by first
perturbing the data with random noise before being released. There is strong interest
in further developing this approach because data curators are often interested
in releasing data to the public without knowledge of how their data will
be used or analyzed. In this scenario, it can be impossible to apply the direct or
query method because they require knowing before hand how the data will be analyzed.
For a recent example of a high profile application of the dissemination approach
see the the U.S. Census Bureau's `TopDown` algorithm [@TopDown]. [@Ji2014]
provide a nice survey of the direct and query based methods applied to 
common machine learning algorithms.

Correctly carrying out statistical inference in the dissemination setting
requires adjusting statistical workflows to account for privacy noise.
As an example, one instance of implementing the dissemination approach for tabular data involves directly
adding independent, random error to each cell. For regression models,
this corresponds to having data with measurement errors
in the covariates. This, unfortunately, violates the assumptions of most statistical models. 
In the presence of such errors, standard estimators can exhibit significant bias [@Gong2022].
Therefore, fitting models without accounting for the added privacy noise can lead to incorrect inference.

Unfortunately, making the necessary adjustments poses formidable mathematical
challenges, even for seemingly simple models like linear regression. 
The difficulty can be seen from considering the marginal likelihood that results from correctly accounting for the injected
privacy noise. This function is often analytically intractable and as a result,
it is difficult or impossible to apply common maximum likelihood methods
to derive estimators. In particular, the marginal likelihood can involve a complex
integral where it is not possible to even evaluate the likelihood
at a point. This makes gradient based methods, like the Newton-Raphson method,
impractical. And approximating the likelihood can be computationally
unfeasible since the integral is also of high dimension. 
Few tools are available to researchers to address these issues,
and their absence is a serious barrier to the wider adoption
of dissemination methods.

The DAPPER package provides a tool for conducting
valid statistical inference in the presence of privacy noise.
It implements the Bayesian framework proposed in [@Ju2022]. This framework describes how to modify
an existing Bayesian model to account for privacy noise. DAPPER
serves as user-friendly interface for implementing the framework in R. It allows the
user to specify a sampler from an existing Bayesian model and
automatically constructing a valid posterior sampler that accounts for the 
added privacy noise.

In the previously mentioned case where privacy noise is
directly added to each cell of a data table, there exist
readily available methods from measurement error literature for making adjustments.
For textbook length treatments on the topic see [@Yi2017; @Carroll2006].
Work in this area mostly focuses on methods which do not require fully specifying the 
measurement error model, since this is often assumed unknown.
However, in differential privacy, the measurement error model is exactly known.
This difference, makes feasible some ideas which the measurement
error community has not previously considered [@Smith2011; @Karwa2015].

The rest of this article is organized as follows. Section 2 covers the necessary background to understand the mathematical notation
and ideas used throughout the paper. Section 3 goes over the main algorithm without
going into mathematical detail, for specifics see [@Ju2022]. Section 4 provides
an overview of the dapper package and discusses important implementation details.
Section 5 contains two example of how one might use the package to analyze the 
impact of adding noise for privacy.

# Background

Let $x = (x_1, \ldots, x_n) \in \mathcal{X}^n$ represent a confidential 
database containing $n$ records. Usually the goal of collecting data 
is to learn some characteristic about the underlying population. 
To accomplish this task, a common approach is to assume the population
is represented by some statistical model $f( \cdot \mid \theta)$. It is often the case that
some function of $\theta$ has relevant meaning to the scientific question at hand. In this setting,
learning characteristics of a population reduces to learning about $\theta$. 

In the Bayesian framework, this is accomplished by drawing samples from the 
posterior $p(\theta \mid x) \propto f(x \mid \theta) p(\theta)$. For 
large data sets, it is common to work with a summary statistic $s = s(x)$
that has much smaller dimension than the original data. Doing so can
greatly simplify calculations. In general, there can be information
loss with using summary statistics, but for models with a sufficient 
statistic, there is no loss. In the context of privacy, providing
a summary statistic can offer a level of anonymity.


## Differential Privacy
While a summary statistic can already anonymize the data, it is still 
possible to deduce information about an individual entry depending on
the distribution of $x$. For additional anonymity, one idea is to 
add noise to the summary statistic $s$. More formally we write 
$s_{dp} \sim \eta(\cdot \mid x)$. Here, $s_{dp}$ is the noise infused version of $s$ and
$\eta$ is a known noise infusion process. The privacy mechanism
$\eta$ is said to be $\epsilon$-differentially private [@Dwork2013] if for all values of
$s_{dp}$, and all "neighboring" databases $(x,x') \in \mathcal{X}^n \times \mathcal{X}^n$ differing
by one record (denoated by $d(x,x')$), the probability ratio is bounded:
\[
\dfrac{\eta(s_{dp} \mid x)}{\eta(s_{dp} \mid x')} \leq \exp(\epsilon), \quad \epsilon > 0.
\]
The parameter $\epsilon$ is called the privacy loss budget, and controls how informative $s_{dp}$
is about $x$. Larger values of $\epsilon$ correspond to less noise being added. 

## Data Augmentation
The idea behind data augmentation is to run a Gibbs sampler on a coupling of two random variables where one 
of the marginals is the target distribution. Suppose we wish to sample from a density $f_{A}$ which is difficult. 
The data augmentation method instead considers sampling from a joint distribution $f(a,b)$. Since
we are ultimately interested in samples from $f_A$, the joint
distribution should be chosen so that (i) the marginal distribution
with respects to $a$ is $f_{A}$ and (ii) $f(a \mid b)$ and $f(b \mid a)$
are easy to sample from. The choice $f$ is not unique and can require
some foresight.

# Methodology
Given data $s_{dp}$, the goal of Bayesian inference is to sample from the 
posterior distribution $p(\theta \mid s_{dp})$. Since the observed likelihood,
$p(s_{dp} \mid \theta)$ often has no simple closed form expression, most standard
sampling schemes do not apply. To conduct privacy aware Bayesian inference, the dapper package implements
the data augmentation algorithm which allows us to sample from $p(\theta \mid s_{dp})$
without needing to specify $p(s_{dp} \mid \theta)$.

The algorithm considers the joint distribution $p(\theta, x \mid s_{dp})$ and
alternates sampling from the two distributions

+ $p(\theta \mid x, s_{dp})$
+ $p(x \mid \theta, s_{dp})$

Since $s_{dp}$ is derived from $x$, we have $p(\theta \mid x, s_{dp}) = p(\theta \mid x)$ which
is just the usual posterior distribution given the confidential data $x$. The dapper
package assumes the user has access to a sampler for $p(\theta \mid x)$. This can
come from any R package such as fmcmc. For the second distribution, $p(x \mid \theta, s_{dp})$, may
only be known up to a constant. The dapper package samples from this distribution by
running a Gibbs like sampler. Each of the $n$ components of $x$ is individually
updated. However unlike the standard Gibbs sampler, each component is updated
using a Metropolis-Hasting algorithm. This method is sometimes called the Metropolis within Gibbs sampler [@Robert2004].

In some cases, sampling from $p(x \mid \theta, s_{dp})$ can be made more efficient
when the summary statistic can be written as the sum of individual
contributions from each observation. More precisely, we say a statistic satisfies
the record additivity property if $\eta(s_{dp} \mid x) = g(s_{dp}, \sum_{i=1}^{n}t_i(x_i, s_{dp}))$
for some known and tractable functions $g, t_1, \ldots, t_n$. 


The algorithm is in the following pseudo code:

1. Sample $\theta^{t+1}$ from $p(\cdot \mid x^{(t)})$.
2. Sample from $p(x \mid \theta, s_{dp})$ using a three step process
    + Propose $x_{i}^{*} \sim f(\cdot \mid \theta)$.
    + If $s$ satisfies the record additive property then
    update $s(x^*, s_{dp}) = t(x,s_{dp}) - t_i(x_i,s_{dp}) + t_{i}(x_i^*, s_{dp})$.
    + Accept the proposed state with probability $\alpha(x_i^* \mid x_i, x_{-i}, \theta)$
    given by:
    
    \[
      \alpha(x_i^* \mid x_i, x_{-i}, \theta) = \min \left\{ 1, \dfrac{\eta(s_{dp} \mid x_i^*, x_{-i})}{\eta(s_{dp} \mid x_i, x_{-i})} \right\}.
    \]


# The Structure of DAPPER

The package is structured around the two functions `dapper_sample` and
`new_privacy`. The first function is used to draw samples from the 
posterior. The second function is used to create a privacy data model object
which the `dapper_sample` function requires as input. The purpose of the data model
object is to collect all the components specific to the data augmentation algorithm
into one bundle. This way, the other arguments into `dapper_sample` pertain only 
to sampling parameters such as the number of iterations.


Since the input to these functions are R functions, there is a great deal of freedom
left up to user. The next two sections describe in detail the inputs into
these functions and highlight some considerations that should be taken
into account in order avoid slow or unexpected behavior.

Before delving into  the specifics of each component, it is necessary to clearly
define how the confidential data is represented. Internally, the 
confidential database is encoded as a 2D matrix. There are often multiple ways
of doing this. For example, if our data consist 
of 100 responses from a two question, yes/no, survey. Then we can either encode
the data as a $2 \times 2$ matrix, or a $100 \times 2$ matrix. Both are mathematically
equivalent, but the $2 \times 2$ matrix will be much more memory efficient.
In general, the representation that uses the least amount of memory should be
used. Correctly specifying the privacy model will require a consistent 
representation among all components.

## Privacy Model

Creating a privacy model is done using the `new_privacy` constructor. The 
main arguments consist of the four components as outlined in the methodology
section.

```{r, echo=TRUE, eval=FALSE}
new_privacy(post_smpl = NULL, lik_smpl = NULL, ll_priv_mech = NULL,
            st_calc = NULL, add = FALSE, npar = NULL)
```

The internal implementation of the DA algorithm in `dapper_sample` requires
some care in how each component is constructed. 

* `lik_smpl` is an R function that samples from the likelihood. Its
syntax should be `lik_smpl(theta)` where `theta` is a vector
representing the likelihood model parameters being estimated. This function
must work with the supplied initial parameter provide in the `init_par`
argument of `dapper_sample` function. The output should be a $n \times k$ matrix
where $k$ is the dimension of the vector of likelihood parameters.


* `post_smpl` is a function which represents the posterior sampler. It should
have the call signature `post_smpl(dmat, theta)`. Where `dmat` is the 
complete data. This sampler can be generated by wrapping mcmc samplers generated from other R packages 
(e.g. \CRANpkg{rstan}, \CRANpkg{fmcmc}, \CRANpkg{adaptMCMC}).
If using this approach, it is recommended to avoid using packages such as \CRANpkg{mcmc}
whose implementation clashes with `gdp_sample`. In the case of \CRANpkg{mcmc},
the Metropolis-Hastings loop is implemented in C which incurs a very large overhead
in `gdp_sample` since it is reinitialized every iteration. In general, repeatedly calling
an R function that hooks into C code is slow. (NOT QUITE ACCURATE FIX LATER)

* `ll_priv_mech` is an R function that represents the log-likelihood of 
$\eta(s_{sdp} \mid x)$. The function can output the log likelihood 
up to an additive constant.

* `st_calc` is an R function which calculates the summary statistic. The optional
argument `add` is a flag which represents whether $T$ is additive or not.


## Sampling

The main function in \pkg{DAPPER} is the `dapper_sample` function. The syntax of the function is:

```{r, echo = TRUE, eval = FALSE}
dapper_sample(data_model, sdp, nobs, init_par, niter = 2000, warmup = floor(niter / 2),
           chains = 1, varnames = NULL)
```

The three required inputs into `dapper_sample` function are the privacy model (`data_model`), the value
of the  observed privatized statistic (`sdp`), and the total number of observations 
in the complete data (`nobs`). The dapper
package is best suited for problems where the complete data can be represented in
tabular form. This is because internally, it is represented as a matrix.

The optional arguments are the number of mcmc draws (`niter`), the 
burn in period (`warmup`), number of chains (`chains`) and character
vector that names the parameters. Running multiple chains can be done in parallel
using the \CRANpkg{furrr} package. Additionally, progress can be monitored
using the \CRANpkg{progressr} package.


The `data_model` input is a `privacy`
object that can be constructed using the `new_privacy` constructor. The
process of constructing a `privacy` object will be discussed in the next section.




# Examples

## 2x2 Contingency Table
A common procedure when analyzing contingency tables is to estimate the 
odds ratio. Something something about safetab to connect back to DP (dont forget citation!). 
As a demonstration, we analyze the UC Berkeley admissions data, which is often
used as an illustrative example of Simpson's paradox. The question is whether
the data suggest there is bias against females during the college admissions 
process. Below is a table of the aggregate admissions result from six departments based on sex.


```{r, echo = FALSE}
set.seed(1)
tmp <- apply(UCBAdmissions, 3, identity, simplify=FALSE)
adm_cnf <- Reduce('+', tmp)
adm_prv <- round(adm_cnf + rnorm(4, mean = 0, sd = 100))
```


```{r, echo = FALSE}
kbl(list(adm_cnf, adm_prv), booktabs = TRUE) %>%
  kable_styling(position = 'center', latex_options = c("hold_position"))
```



```{r, echo = FALSE}
#generate 2x2 table data
x <- c(adm_cnf)
sdp <- c(adm_prv)
```

Below we walk through the process of defining a privacy model.


1. `lik_smpl`: Conditional on the table total, the table counts follow a multinomial
distribution. We can easily draw from this distribution using the
`rmultinom` function in the `base` stats package. Note, in this example,
the return value of one sample from `rmultinom` is a $4 \times 1$ matrix. In order to 
conform with `dapper_sample` we must convert the matrix to a vector.
    ```{r, echo = TRUE}
    lik_smpl <- function(theta) {
      t(rmultinom(1, 4526, theta))
    }
    ```
2. `post_smpl`: Given confidential data $X$ we can derive the posterior analytically
using a Dirichlet prior. In this example, we use a flat prior which
corresponds to Dirch(1) distribution. A sample from the Dirichlet distribution
can be generated using the gamma distribution via the following relation (INSERT)
    ```{r, echo = TRUE}
    post_smpl <- function(dmat, theta) {
      x <- c(dmat)
      t1 <- rgamma(length(theta), x + 1, 1)
      t1/sum(t1)
    }
    ```
3. `st_calc`: The complete data can be represented in two ways. Micro vs cell totals.
(what section to introduce?) This function must return a vector.
    ```{r, echo = TRUE}
    st_calc <- function(dmat) {
      c(dmat)
    }
    ```
4. `ll_priv_mech`: Privacy Mechanism
Guassian white noise is added to each cell total. Hence given
confidential data $(n_{11}, n_{22}, n_{12}, n_{21})$
\[
\eta(s_{dp} \mid x) = \prod \phi(s_{sd}; n_{ij}, 100^2)
\]
    ```{r, echo = TRUE}
    ll_priv_mech <- function(sdp, x) {
      dnorm(sdp - x, mean = 0, sd = 100, log = TRUE)
    }
    ```


<!-- end of list -->

Once privacy model has been defined we can run `gdp_sample`

```{r, echo = TRUE}
library(DPloglin)
dmod <- new_privacy(post_smpl = post_smpl,
                    lik_smpl = lik_smpl,
                    ll_priv_mech = ll_priv_mech,
                    st_calc = st_calc,
                    add = FALSE,
                    npar = 4,
                    varnames = c("pi_11", "pi_21", "pi_12", "pi_22"))
                  
dp_out <- dapper_sample(dmod,
                  sdp = c(adm_prv),
                  niter = 10000,
                  warmup = 1000,
                  chains = 1,
                  init_par = rep(.25,4))
```

results can be quickly summarized using the `summary` function

```{r,echo = FALSE}
summary(dp_out)
```

Diagnostic checks can be done using the \pkg{Bayesplot} package.

```{r, fig.height=3, fig.width=5, fig.align='center'}
plot(dp_out)
```

log odds distribution

```{r, fig.height=3, fig.width=5, fig.align='center'}
tv <- dp_out$chain
or <- as.numeric((tv[,1] * tv[,4]) / (tv[,2] * tv[,3]))
quantile(or, c(.025, .50, .975))

ggplot(tibble(x=or), aes(x)) + geom_histogram() + xlim(-1,10)
```
(Insert gdp Analysis Here)

For clean data, a estimate for the odds ratio and a confidence interval
can be constructed using Woolf's method (i.e. A wald confidence interval).
It uses the fact that the log of the odds ratio is approximately 
normal for large sample sizes.
\[
\log\left(\dfrac{n_{11} \cdot n_{22}}{n_{12} \cdot n_{21}}\right) 
  \pm z_{\alpha/2}\sqrt{\dfrac{1}{n_{11}} + \dfrac{1}{n_{22}} + \dfrac{1}{n_{12}} + \dfrac{1}{n_{21}}}
\]

```{r, echo = TRUE}
or_confint <- function(x, alpha) {
  or <- log(x[1] * x[4]/ (x[2] * x[3]))
  se <- sqrt(sum(1/x))
  c(or - qnorm(alpha/2) * se, or + qnorm(alpha/2) * se)
}

#clean data
exp(or_confint(x, .95))

#privitized data
exp(or_confint(sdp, .95))
```

## Linear Regression

## Logistic Regression

```{r, eval = FALSE, include= FALSE}


```

# Summary

This package is cool. You should install it.